PLUNGER: Reproducing FLUSH+RELOAD: A High-Resolution, Low-Latency Side Channel Attack On GnuPG

### 2014-12-17 paper/paper.tex
Some minor corrections

    In their paper, Yarom and Falkner\cite{YF13} presented a side channel
    attack, which they named \flushreload{} on the L3 caches of Intel x86
    processors. The attack takes advantage of content-based memory page sharing
    between non-trusting processes to monitor memory accesses of the victim
    process. This technique can be used on GnuPG pre-1.4.14 to identify the
    binary sequence of a RSA private exponent by discerning which branches are
    taken.

@@

A typical cache architecture is laid out in three levels, generally referred to
as the \textit{L1}, \textit{L2}, and \textit{L3}. The L3 cache is usually the
largest of the three, then the L2, and finally the L1; L1 and L2 caches are also
generally allocated on a per-core basis in the case of multi-core CPUs, whereas
the L3 cache is usually shared between all cores, as it is the largest. L1
caches are also typically divided into \textit{instruction} sections and
\textit{data} sections, each reserved for that particular type of instruction.
It is equally important for instruction memory as well as data memory to be
cached, as both require fast access in order for the CPU to be performant. For
example, the Intel Haswell i7 4600U CPU has a 128KiB L1 cache, divided into two
64KiB caches for instructions and data; a 512KiB L2 cache, and a 4MiB L3 cache.

Naturally, the closer a cache is to a core, the lower the latency in accessing
that cache by the core, as there are fewer buses to traverse.  On Intel
architectures, the caches are also \textit{inclusive} in their data, so as to
aid in cache coherence, i.e.  ensuring data coherence in the cache across cores.
Cache inclusivity means that whatever data is contained in a core's L1 cache
will also be contained in the core's L2 cache as well as the CPU's L3 cache.
Therefore, when data gets evicted from the L3 cache, it will also be evicted
from the L2 and L1 caches.

@@

Modern operating systems also implement a technique known as \textit{page
sharing}, which helps to reduce the memory footprint of running processes by
avoiding replicating pages of memory that are already allocated. Page sharing
may also be used as an efficient, safe channel between two co-operating
processes.

Windows, Linux, and possibly OS X \footnote{It is unclear whether or not OS X
implements page deduplication, but as far as we can tell it does} also perform a
more aggressive form of page sharing known as \textit{page deduplication}. With
this scheme, the operating system scans memory pages during idle CPU cycles, and
looks for pages that are identical in content. It then collapses them into one
singular page that is then shared by all processes that are accessing the
content of that particular page. As a result of this page deduplication scheme,
processes may be unknowingly or unintentionally sharing memory with each other.
However, to prevent malicious access to the shared pages, the operating system
maps shared pages as copy-on-write; that is, when a process wishes to write to a
page, it must first make a local copy for itself that it can safely modify. When
a write is requested, the CPU traps to the kernel, which then remaps a copy of
the desired page for the process requesting the write, and then returns control
to the calling process. Read operations in this scheme are safe, as they do not
modify the targeted page.

@@

RSA is a standard algorithm for public key encryption. The scheme works as
follows. Choose two distinct prime numbers, $p$ and $q$.

\begin{enumerate}
    \item Compute $N = pq$.
    \item Compute $\phi(N) = \phi(p)\phi(q) = (p-1)(q-1)$
    \item Choose $e, 1 \leq e \leq \phi(N)$, where $gcd(e,\phi(N)) = 1$.
    \item Compute $d = e^{-1}\ mod\ \phi(N)$
    \item Let the private key be (d,N) and the public key be (e,N)
\end{enumerate}

@@

Of interest is the algorithm used to compute $c^d\ mod\ N$. This is known as the
problem of modular exponentiation. The most straightforward method of performing
modular exponentiation is to reduce modulo $N$ after each multiplication. This
would then take $O(d)$ multiplications. However, there exists a much more
efficient algorithm known as Square-and-Multiply\cite{Knuth:1997:ACP:270146}.
This method is derived from the binary representation of the exponent $d$.
Consider $d = \sum_{i=0}^{n-1}x_i 2^i$. Then $c^d = c^{(\sum_{i=0}^{n-1}x_i
2^i)} = \prod_{i=0}^{n-1}c^{x_i 2^i}$. Then reducing modulo $N$ we get obtain
$m$ as desired. The following code shows GnuPG 1.4.13's implementation of
Square-and-Multiply. We can proceed to implement the algorithm by scanning the
binary representation of $d$ from right to left. If the bit is $1$, we will
multiply and then square, and if the bit is a $0$, we will just square.
Inserting reduces to keep memory use manageble, we see that when the bit is $1$,
we will multiply, reduce, square, and reduce, and when the bit is $0$, we will
square, and then reduce.

@@

GnuPG (Gnu Privacy Guard) is a free implementation of PGP. As such, GnuPG allows
for encryption, signing, and decryption of messages. It implements several
public key algorithms, including DSA, ElGamal, and importantly for our purposes,
RSA. As mentioned above, in particular, the GnuPG implementation of RSA uses
Square-and-Multiply for modular exponentiation. As of GnuPG version 1.4.14, the
if switch guarding the multiply and reduce no longer prevents that branch from
being taken, rather, in the case of the bit being a $0$, it will simply discard
the results of the multiply in the reduce. This was a consequence of the
publication of the \flushreload{} attack by Yarom and Falkner\cite{YF13} that we
will describe momentarily.

@@

The \flushreload{} attack takes advantage of page deduplication on modern
processors. The basic attack works as follows. The attacker loads into memory
some block of data that it believes the victim to also be accessing. The system
then deduplicates these two identical pages. The attacker can then use the
instruction \texttt{clflush} to flush this from its $L1$ cache, which flushes it
from the $L3$ cache, and consequently the $L1$ cache of the victim as well. The
attacker then waits for some time slot $T$, and attempts to access a cache line
in this page, and measures the time (in cycles) it takes for the line to be
loaded. If the cache line was accessed by the victim during $T$, the time taken
will be shorter, and if it was not accessed, the time taken will be longer.
Consequently, the attacker can determine if the line was accessed during $T$.

@@

Our implementation of the \flushreload{} technique is an adaptation of the
implementation as described in Yarom and Falkner\cite{YF13}. In the original
implementation, the \texttt{probe} function measures the time taken to access a
given address, immediately flushes the corresponding cache line, and returns
whether the cycles elapsed stays within a fixed threshold.  We make a minor
modification to the \texttt{probe} function---returning the number of cycles
elapsed instead of the result of the thresholding---to allow us to obtain data
that can be easily checked visually.

@@

Leaving debug symbols in the executable made it easier to map the relevant GPG
source code lines to corresponding addresses of the instructions in the binary's
.text section. GPG performance is not affected because these symbols are removed
at runtime. Finding the relevant instructions is still possible in a stripped
GPG executable, though it will require some reverse engineering.

@@

Unfortunately, while the Python script that we developed to interpret these
results was able to recover many of the bits in $d_p$ and $d_q$, there were too
many errors in recovery to reliably reconstruct the original private RSA
exponent.

@@

We were unable to extract any binary strings from the OS X results. However,
they looked promising, and we believe that page deduplication is happening
properly, so with further experimentation, the modulus might be extracted. There
are a few reasons we believe this to be the case. First, on many runs the
proportions of square, reduce, and multiplies were correct. A random assortment
of bits in the modulus combined with the access patterns of the algorithm should
produce a particular ratio, roughly $4:2:1$ reduces to squares to multiplies,
and we did observe this. Furthermore, by tweaking the probe placement and
timing, the number of accesses observed per slot was between 1 and 2 on average,
which is close to the ideal of exactly one access per slot. Lastly, the banding
by cache level was very evident, similar to the EC2 results above. There was
little ambiguity about which probes resulted in cache hits and which did not.

As of writing, we did not examine the assembly of the GnuPG binary to optimally
place probes, rather, we used the method of setting breakpoints in GDB by line
number and observing the addresses on which they were set. It is possible that
switching methods would have produced clean enough results to retrieve the
mdodulus.

@@

Arch Linux provided the poorest results by far. No meaningful data was
discernible under any configuration, as there is timing noise for the entire
duration of spying, which in the case of the figure below, occurred roughly
between timeslots 7000 and 18000 on the x-axis.. Furthermore, it was fairly
opaque as to why this was occurring, as there were no significant differences
between this and the configuration running on OS X. The largest obvious
difference between the two systems is that GnuPG on Arch Linux was compiled
under GCC 4.9.2, as opposed to GCC 4.2 on OS X; however, it is not clear how
this would change the output this drastically, or if it did at all. Two
additional hypotheses for why this behaviour was exhibited were that Arch Linux
was being run as a guest OS on a Windows 8.1 host, and that it was being run on
a Haswell-series Intel i7, which might require compiling the GnuPG binary
differently and therefore might exhibit different behaviour, as this CPU series
did not exist at the time of the original paper's publishing.  However, both of
these hypotheses were disproven by testing on a second machine running Arch
Linux as a host OS on an older Intel Celeron CPU, in which the attack produced
the same noisy results.

@@

Our results were quite sensitive to probe placement. The ideal probing addresses
are the ones that are always executed when the targeted operation
(\texttt{mpih\_sqr\_n\_basecase}, \texttt{mpihelp\_divrem}, and
\texttt{mpihelp\_mul\_karatsuba\_case}) is executed. Finding these addresses,
however, was not as straightforward as we thought it would be.

Since GPG is built with \texttt{-O2} optimzations, the assembly output does not
map one-to-one with the lines of source code; optimizations tend to do a
considerable amount of assembly ``jumbling'' (duplication, rearrangement, and
inlining of code). As a result, we found that setting probes to within
frequently accessed addresses, such as within loop bodies as suggested by Yarom
and Falkner\cite{YF13}, did not work as well as expected.

Many of our preliminary results were plagued by a lot of noise, a lot of which
resulted from speculative execution. To reduce the effects of this, we attempted
to probe the very close to the ends of the respective functions. We found that
simply setting breakpoints to the relevant lines in GDB would not yield good
addresses to probe. Even setting a breakpoint to the last line of the function
in the source would not return an address of an instruction close to the return.
To find the best addresses, we disassembled the GPG binary and located addresses
near each function's \texttt{retq} instruction.

@@

@@

We successfully reproduced parts of the \flushreload{} attack. While results
were mixed on various platforms, we found moderate success on EC2, which, in
particular, is an interesting environment vulnerable to this type of attack
because of its shared hardware access through a VM. Though GnuPG has fixed this
vulnerability starting at version 1.4.14, this shows that other binaries running
on EC2 may prove vulnerable.

We have endeavored to provide more details on some parts of the implementation
of this attack so as to reduce the barrier to future reproductions and
extensions of these results. Additionally, full source code is available upon
request.

\section{Future Work} % David

A possible future application of the \flushreload{} technique would be to
perform this attack across virtual machines. That is, using the attack to steal
data from one virtual machine whilst running the attack on another. Cross-VM
attacks are described in further detail by Apecechea et al.
\cite{cryptoeprint:2014:248}; the basic principle behind the attack would be to
detect and spy upon a victim, i.e. a user on another VM, running GnuPG (or some
equivalent cryptographic process) and steal the keys used in the process.  Such
an attack clearly has its appeal-- it would expose a major vulnerability in
virtual machines, which are used in countless modern software applications.

>>>>

    In their paper, Yarom and Falkner\cite{YF13} presented a side channel
    attack, named \flushreload{}, on the L3 caches of Intel x86 processors. The
    attack takes advantage of content-based memory page sharing between
    non-trusting processes to monitor memory accesses of a victim process. This
    technique can be used on GnuPG pre-1.4.14 to identify the binary sequence of
    an RSA private exponent by discerning which branches are taken.

@@

A typical cache architecture is laid out in three levels, generally referred to
as the \textit{L1}, \textit{L2}, and \textit{L3}. The L3 cache is usually larger
than than the L2, which itself is usually larger than the L1; L1 and L2 caches
are also generally allocated on a per-core basis in the case of multi-core CPUs,
whereas the L3 cache is usually shared between all cores.  L1 caches are also
typically divided into \textit{instruction} sections and \textit{data} sections,
each reserved for that particular type of instruction.  It is equally important
for instruction memory as well as data memory to be cached, as both require fast
access in order for the CPU to be performant. For example, the Intel Haswell i7
4600U CPU has a 128KiB L1 cache, divided into two 64KiB caches for instructions
and data; a 512KiB L2 cache, and a 4MiB L3 cache.

Naturally, a cache closer to a core has a lower access latency because the core
has to traverse fewer buses. On Intel architectures, the caches are considered
\textit{inclusive}, meaning that whatever data is contained in a core's L1 cache
will also be contained in the core's L2 cache and the CPU's L3 cache.
Therefore, when data are evicted from the L3 cache, they will also be evicted
from the L2 and L1 caches.

@@

Modern operating systems also implement a technique known as \textit{page
sharing}, which helps reduce the memory footprint of running processes by
avoiding redundant replication of memory pages. Page sharing may also be used as
an efficient, safe channel between two co-operating processes.

Windows, Linux, and possibly OS X\footnote{It is unclear whether or not OS X
implements page deduplication, but as far as we can tell it does} also perform a
more aggressive form of page sharing known as \textit{page deduplication}. With
this scheme, the operating system scans memory pages during idle CPU cycles, and
looks for pages that are identical in content. It collapses those pages into a
single page that is then shared by all processes that were accessing those
identical pages. As a result of this page deduplication scheme,
processes may be unknowingly or unintentionally sharing memory with one another.
However, to prevent malicious access to the shared pages, the operating system
maps shared pages as copy-on-write; that is, when a process wishes to write to a
page, it must first make a local copy for itself that it can safely modify. When
a write is requested, the CPU traps to the kernel, which then remaps a copy of
the desired page for the process requesting the write, and then returns control
to the calling process. Read operations in this scheme are safe, as they do not
modify the targeted page.

@@

RSA is a standard cryptosystem for public key encryption. The scheme works as
follows. Choose two distinct prime numbers, $p$ and $q$.

\begin{enumerate}
    \item Compute $N = pq$.
    \item Compute $\phi(N) = \phi(p)\phi(q) = (p-1)(q-1)$
    \item Choose $e, 1 \leq e \leq \phi(N)$, where $gcd(e,\phi(N)) = 1$.
    \item Compute $d = e^{-1}\ mod\ \phi(N)$
    \item Let the private key be $(d,N)$ and the public key be $(e,N)$
\end{enumerate}

@@

Of interest is the algorithm used to compute $c^d\ mod\ N$. This is known as the
problem of modular exponentiation. The most straightforward method of performing
modular exponentiation is to reduce modulo $N$ after each multiplication. This
would then take $O(d)$ multiplications. However, there exists a much more
efficient algorithm known as Square-and-Multiply\cite{Knuth:1997:ACP:270146}.
This method is derived from the binary representation of the exponent $d$.
Consider $d = \sum_{i=0}^{n-1}x_i 2^i$. Then $c^d = c^{(\sum_{i=0}^{n-1}x_i
2^i)} = \prod_{i=0}^{n-1}c^{x_i 2^i}$. Then reducing modulo $N$ we obtain
$m$ as desired. The following code shows GnuPG 1.4.13's implementation of
Square-and-Multiply. We can proceed to implement the algorithm by scanning the
binary representation of $d$ from right to left. If the bit is $1$, we will
multiply and then square, and if the bit is a $0$, we will just square.
Inserting reduces to keep memory use manageble, we see that when the bit is $1$,
we will square, reduce, multiple, and reduce, and when the bit is $0$, we will
only square and reduce.

@@

GnuPG (GNU Privacy Guard) is a free implementation of PGP. As such, GnuPG allows
for encryption, signing, and decryption of messages. It implements several
public key algorithms, including DSA, ElGamal, and importantly for our purposes,
RSA. As mentioned above, in particular, the GnuPG implementation of RSA uses
Square-and-Multiply for modular exponentiation. As of GnuPG version 1.4.14, the
\texttt{if} branch guarding the multiply and reduce no longer prevents that branch from
being taken; rather, in the case of the bit being a $0$, it will simply discard
the results of the multiply in the reduce. This was a consequence of the
publication of the \flushreload{} attack by Yarom and Falkner\cite{YF13} that we
will describe momentarily.

@@

The \flushreload{} attack takes advantage of page deduplication on modern
processors. The basic attack works as follows. The attacker loads into memory
some block of data that it believes the victim to also be accessing. The system
then deduplicates these two identical pages. The attacker can then use the
instruction \texttt{clflush} to flush memory lines in that block from the CPU caches. The
attacker then waits for some time $T$ and attempts to access a memory line
in this page, and measures the time (in cycles) it takes for the line to be
loaded. If the memory line was accessed by the victim during $T$, the time taken
will be shorter because it will be cached, and if it was not accessed, the time taken will be longer.
Consequently, the attacker can determine if the line was accessed during $T$.

@@

Our implementation of the \flushreload{} technique is an adaptation of the
implementation as described in Yarom and Falkner\cite{YF13}. In the original
implementation, the \texttt{probe} function measures the time taken to access a
given address, immediately flushes the corresponding memory line from the cache, and returns
whether the cycles elapsed stays within a fixed threshold.  We make a minor
modification to the \texttt{probe} function---returning the number of cycles
elapsed instead of the result of the thresholding---to allow us to obtain data
that can be easily checked visually.

@@

Leaving debug symbols in the executable made it easier to map the relevant GPG
source code lines to the addresses of the corresponding instructions in the binary's
.text section. GPG performance is not affected because these symbols are removed
at runtime. Finding the relevant instructions is still possible in a stripped
GPG executable, though it will require some reverse engineering.

@@

@@

We were unable to extract any binary strings from the OS X results. However,
the output timing data looked promising and leads us to believe that OS X does
implement page deduplication. On many runs the
proportions of square, reduce, and multiplies were correct. A random assortment
of bits in the modulus combined with the access patterns of the algorithm should
produce a particular ratio, roughly $4:2:1$ reduces to squares to multiplies,
which we did observe. Furthermore, by tweaking the probe placement and
timing, the number of accesses observed per slot was between 1 and 2 on average,
which is close to the ideal of exactly one access per slot. Lastly, the banding
by cache level was very evident, similar to the EC2 results above. There was
little ambiguity about which probes resulted in cache hits and which did not.

As of writing, we did not examine the assembly of the GnuPG binary on OS X for
optimal probe placement; rather, we used the method of setting breakpoints in GDB by line
number and observing the addresses on which they were set. It is possible that
switching methods would produce better results.

@@

Arch Linux provided the poorest results by far. No meaningful data were
discernible under any configuration because there was too much timing noise
during the spying, which, in the case of the Figure~\ref{fig:data-arch}, occurred roughly
between timeslots 7000 and 18000 on the x-axis. Given that there were no
significant differences between this and the configuration running on OS X, we
were unsure why this kept occurring. The most obvious
difference between the two systems is that GnuPG on Arch Linux was compiled
under GCC 4.9.2 instead of GCC 4.2, as it was on OS X; however, it was not clear how
this would change the output so drastically, or if it did at all. Our two
hypotheses for this behaviour were that Arch Linux
was being run as a guest OS on a Windows 8.1 host, and that it was being run on
a Haswell-series Intel i7, which could have affected GCC's compilation of GnuPG
because this CPU series
did not exist at the publication time of the original paper.  However, both of
these hypotheses were disproven by testing on a second machine running Arch
Linux as a host OS on an older Intel Celeron CPU, in which the attack produced
the same noisy results.

@@

Our results were quite sensitive to probe placement. The ideal probing addresses
are the ones that are always executed when the targeted operations
(\texttt{mpih\_sqr\_n\_basecase}, \texttt{mpihelp\_divrem}, and
\texttt{mpihelp\_mul\_karatsuba\_case}) are executed. Finding these addresses,
however, was not as straightforward as we thought it would be.

Since GPG is built with \texttt{-O2} optimzations, the assembly output does not
map one-to-one with the lines of source code; optimizations tend to do a
considerable amount of assembly ``jumbling'' (duplication, rearrangement, and
inlining of code). As a result, we found that setting probes to within
frequently accessed addresses, such as within loop bodies as suggested by Yarom
and Falkner\cite{YF13}, did not work as well as we expected.

Many of our preliminary results were plagued by a lot of noise, much of which
resulted from speculative execution. To reduce the effects of this, we attempted
to probe the very close to the ends of the respective functions. We found that
simply setting breakpoints to the relevant lines in GDB would not yield good
addresses to probe. Even setting a breakpoint to the last line of the function
in the source would not return an address of an instruction close to the return.
To find the best addresses, we disassembled the GPG binary using
\texttt{objdump} and located addresses near each function's \texttt{retq}
instruction.

@@

We developed a Python script to analyze the timing data. While the script was
able to recover many of the bits in $d_p$ and $d_q$, there were too many errors
in recovery preventing us from reliably reconstructing the private RSA exponent.
Further work will be needed to improve the robustness of the script.

@@

We successfully reproduced parts of the \flushreload{} attack. While results
were mixed on various platforms, we found moderate success on EC2, which, in
particular, is an interesting environment vulnerable to this type of attack;
it shares hardware access with other guest VMs on the same host, opening the
possibility of a cross-VM attack. Although this vulnerability was fixed in GnuPG
version 1.4.14, this shows that other programs running on EC2 may be susceptible
to this technique.

We have endeavored to provide more details on some parts of the implementation
of this attack to reduce the barrier to future reproductions and
extensions of these results. Additionally, full source code is available upon
request.

\section{Future Work}

A possible future application of the \flushreload{} technique would be to
perform this attack across virtual machines. That is, using the attack to steal
data from one virtual machine whilst running the attack on another. Cross-VM
attacks are described in further detail by Apecechea et al.
\cite{cryptoeprint:2014:248}; the basic principle behind the attack would be to
detect and spy upon a victim, i.e. a user on another VM, running GnuPG (or some
equivalent cryptographic process) and steal the keys used in the process. Such
an attack clearly has its appeal: it would expose a major vulnerability in
virtual machines, which are used in countless modern software applications.
